---
title: "Import and parse data"
author: "Vincent Crabbe (adapted from Indra Bervoets & Amber Bernauw)"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding=encoding, knit_root_dir="..", output_dir="../Output")})
output: 
  html_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(readxl)
library(dplyr)
library(data.table)
source("../helpers.R")
```
I usually organise my R projects as follows:

* One separate folder for each experiment, e.g. 20260101_VCRA_projectX_experiment1.
* Each experiment folder contains an `.Rproj` file at the root and `Data/`, `Script/`, and `Output/` subfolders.

```{r}
current_dir <- getwd()
# print(current_dir)
data_dir <- file.path(current_dir, "Data")
output_dir <- file.path(current_dir, "Output")

project_name <- basename(current_dir)
```
Don't forget to change the file name!

# Read in data

```{r}
OD_values <- read_xlsx(
  file.path(data_dir, "example_data.xlsx"),
  sheet = "Plate 1 - Sheet1",
  range = "B30:CU211" # You can read past the actual data, but make sure that you don't include the table below
)

OD_values <- na.omit(OD_values) # Remove the bottom rows containing NAs (if present) to strip the data frame to the right size
n_timepoints <- nrow(OD_values)

head(OD_values)
```
Ensure that all data is included in the dataframe!

# Read in template design
Sometimes, we don't use the entire microplate. There are two ways to deal with this situation:

* Don't fill in unused rows or columns in the template.csv file. Leave out corresponding measurements in `OD_values`. This strategy only works if the used wells are shaped as a rectangle.
* Label the unused wells as "blank" in the template.csv file. At the end of the script, remove data labelled as "blank".

```{r}
template <- read.csv(
  file.path(data_dir, "template.csv"),
  header = FALSE,
  skip = 1
)
template[1] <- NULL

n_plate_cols <- ncol(template)
n_plate_rows <- nrow(template)
n_plate_wells <- n_plate_rows * n_plate_cols

# Reset the column names (purely aesthetic)
colnames(template) <- seq(1:n_plate_cols)
```

# Building the big data frame
`data` contains the data in long format: one row per well, per time point.
That is, `r n_timepoints * n_plate_wells` rows in total.

## Plate layout
We start by constructing a small data frame `p` containing the plate layout, consisting of `r n_plate_wells` rows.
This information is collected from the 'template.csv' file in the Data folder.

**Every cell of the template file should adhere to following format:**
strain/concentration_of_the_molecule/name_of_the_applied_molecule/replicate_number

```{r}
# Initialise data frame
p <- data.frame(
  row = rep(seq(1, n_plate_rows), each = n_plate_cols), # 1 1 1 ... 1 2 2 2 2 ...
  col = rep(seq(1, n_plate_cols), n_plate_rows) # 1 2 3 4 ... 12 1 2 3 4 ...
)

for (w in 1:n_plate_wells) {
  # Get the matching string from the template (here "row" means plate row)
  template_string <- template[p[w, "row"], p[w, "col"]]

  # The string is split at the slashes so we can extract the different fields
  tmp <- unlist(strsplit(template_string, "/"))
  p[w, "strain"] <- tmp[1]
  p[w, "mol_ID"] <- tmp[2]
  p[w, "mol_conc"] <- tmp[3]
  p[w, "replicate"] <- tmp[4]
}

p$strain <- as.factor(p$strain)
p$mol_ID <- as.factor(p$mol_ID)
p$mol_conc <- as.factor(p$mol_conc)
p$replicate <- as.factor(p$replicate)

p <- reorder_factor_column(p, "mol_conc")
```

## Adding all the information together
Now, we loop over `data` row by row and add the corresponding data from `OD_values` to it.
```{r}
temperatures <- OD_values |>
  select(starts_with("T°")) |>
  pull(1)

# Transfer metadata from `p` to `data`
data <- data.frame(
  cycle = rep(seq(1, n_timepoints), each = n_plate_wells),
  time = rep(format_time(OD_values$Time), each = n_plate_wells),
  well = rep(generate_well_ids(n_plate_rows, n_plate_cols), times = n_timepoints),
  row = rep(p$row, n_timepoints),
  col = rep(p$col, n_timepoints),
  temperature = rep(temperatures, each = n_plate_wells),  # Temperature is only measured once per cycle
  strain = rep(p$strain, n_timepoints),
  mol_ID = rep(p$mol_ID, n_timepoints),
  mol_conc = rep(p$mol_conc, n_timepoints),
  replicate = rep(p$replicate, n_timepoints)
)
head(data)

for (r in 1:nrow(data)) {
  # Where to look in the plate reader data (here "row" means dataframe row):
  row <- data[r, "cycle"]
  col <- data[r, "well"]

  # Add the matching OD values to `data`
  data[r, "OD"] <- as.double(sub(",", ".", OD_values[[row, col]])) # Replace commas with decimal points
}

head(data)
```

# Write and save data file

```{r}
save(data, project_name, file = file.path(output_dir, "01_import.RData"))
fwrite(data, file.path(output_dir, paste(project_name, "processed.csv", sep = "_")))
```
