---
title: "Kinetic growth experiment: example script"
author: "Vincent Crabbe (adapted from Indra Bervoets & Amber Bernauw)"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding=encoding, knit_root_dir="..", output_dir="../Output")})
output: 
  html_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(readxl)
library(dplyr)
library(data.table)
```



```{r}
current_dir <- getwd()
# print(current_dir)
data_dir <- file.path(current_dir, "Data")
output_dir <- file.path(current_dir, "Output")

file_name <- "example_data"
```
Do not forget to change the file name!


```{r include=FALSE}
# Functions

format_time <- function(df) {
  #' Add hours since first time point
  #'
  #' Adds a new column `time_h` to a data frame, representing the number of
  #'   hours passed since the first value in the `Time` column.
  #'
  #' @param df A data frame with a POSIXct column named `Time`.
  #' 
  #' @return A data frame with an additional numeric column `time_h`.
  #'
  #' @examples
  #' df <- data.frame(Time = as.POSIXct(c("2024-01-01 08:00", "2024-01-01 10:30")))
  #' format_time(df)
  #' 
  df$time_h <- as.numeric(difftime(df$Time, df$Time[1], units = "hours"))
  df
}
```

# Read in data

```{r}
OD_values <- read_xlsx(
  file.path(data_dir, paste(file_name, "xlsx", sep = ".")),
  sheet = "Plate 1 - Sheet1",
  range = "B30:CU211" # You can read past the actual data, but make sure that you don't include any fluorescence values
  )

# Remove the bottom rows containing NAs (if present) to strip the data frame to the right size
OD_values <- na.omit(OD_values)

OD_values <- format_time(OD_values)
n_timepoints <- nrow(OD_values)

head(OD_values)
```
Ensure that all data is included in the data frame!

# Read in template design 96-well plate

```{r}
template <- read.csv(
  file.path(data_dir, "template.csv"),
  header = FALSE,
  skip = 1
  )

# These variables will save your butt if you didn't fill the entire plate
n_plate_cols <- ncol(template)
n_plate_rows <- nrow(template)
n_plate_wells <- n_plate_rows * n_plate_cols

# Reset the column names (purely aesthetic)
colnames(template) <- seq(1:n_plate_cols)
```

# Building the big data frame
`data` contains the data in long format: one row per well, per time point.
That is, `r n_timepoints * n_plate_wells` rows in total.

## Plate layout: data frame of 96 rows
We start by constructing a small data frame `p` containing the plate layout, consisting of `r n_plate_wells` rows.
This information is collected from the 'template.csv' file in the Data folder.

**Every cell of the template file should adhere to following format:**
strain/concentration_of_the_molecule/name_of_the_applied_molecule/replicate_number

```{r}
# Initialise data frame
p <- data.frame(
  row = rep(seq(1, n_plate_rows), each = n_plate_cols), # 1 1 1 ... 1 2 2 2 2 ...
  col = rep(seq(1, n_plate_cols), n_plate_rows)         # 1 2 3 4 ... 12 1 2 3 4 ...
)

for (w in 1:n_plate_wells) {
  # Get the matching string from the template (here "row" means plate row).
  template_string <- template[p[w, "row"], p[w, "col"]]
  
  # The string is split at the slashes so we can extract the different fields
  tmp <- unlist(strsplit(template_string, "/"))
  p[w, "strain"] <- tmp[1]
  p[w, "mol_ID"] <- tmp[2]
  p[w, "mol_conc"] <- tmp[3]
  p[w, "replicate"] <- tmp[4]
}

p$strain <- as.factor(p$strain)
p$mol_ID <- as.factor(p$mol_ID)
p$mol_conc <- as.factor(p$mol_conc)
p$replicate <- as.factor(p$replicate)

# Reorder levels of `mol_conc` column according to increasing concentration
sorted_labels <- paste(sort(as.numeric(levels(p$mol_conc))))
p$mol_conc <- factor(p$mol_conc, levels = sorted_labels, ordered = TRUE)
```

## Adding all the information together
Now, we loop over `data` row by row and add the corresponding data from `OD_values` to it.
```{r}
data <- data.frame(
  time = rep(OD_values$time_h, each = n_plate_wells),
  sample = rep(seq(1, n_timepoints), each = n_plate_wells),
  well = rep(seq(1, n_plate_wells), n_timepoints),
  row = rep(p$row, n_timepoints),
  col = rep(p$col, n_timepoints),
  strain = rep(p$strain, n_timepoints),
  mol_ID = rep(p$mol_ID, n_timepoints),
  mol_conc = rep(p$mol_conc, n_timepoints),
  replicate = rep(p$replicate, n_timepoints)
)
head(data)

# Loop over plate reader measurements only. Modify the range if you didn't completely fill the plate!!
OD_values <- select(OD_values, A1:H12)
for (r in 1:nrow(data)) {
  # Where to look in the plate reader data:
  x <- data[r, "well"]
  y <- data[r, "sample"]

  # Add the matching OD values to `data`
  data[r, "OD"] <- as.double(sub(",", ".", OD_values[y, x])) # Replace commas with decimal points
}

head(data)
```

# Write and save data file

```{r}
save(data, file_name, file = file.path(output_dir, "01_import.RData"))
fwrite(data, file.path(output_dir, paste(file_name, "processed.csv", sep = "_")))
```
